Trending assessment for nowcasts, measurements, and forecasts

### 2024-03-07 sections/01-introduction.tex
Update on Overleaf.

\unsure[inline]{Der Beginn ist jetzt sehr auf Measurement fixiert. Könnte man das vielleicht umdrehen: Analysen von Measurement/Nowcast/Forecast sind im Moment vor allem bezüglich Abstand (RMSE; MAE; aber auch probabilistische Analyse). Wichtige Info ist aber auch auch, wie gut ein Anstieg/Fall vorhergesagt wird. Dann drauf eingehen, wie das bisher in den Bereichen gemacht wird.}

@@

\textbf{Contributions.} The main contributions of this paper are summarized below
\begin{itemize}

\item We formalise the concept of trending and show how the predicted changes can be meaningfully computed to assess trendability.

\item We present different variants of trending measures that consider noiseless data as well as noisy and small non-informative changes.

\item We introduce the conditional trending plot, a new graphical method for assessing local trends.

\item We introduce bootstrap methods for calculating  confidence intervals for the presented treding ratios.

\item We extend the concept of trending to probabilistic predictions.

\item We apply trending evaluation to three different applications: Measurement, nowcasting and prediction data.

\item We provide a ready-to-use code for trending detection. The code and the source code to replicate the results of our study are available in our public repository at ... .
\end{itemize}

The remainder of this paper is structured as follows. 
In Section \ref{sec:trending}, we formalize the concept of trending independently of the application domain and investigate quantification approaches for noise-free and noisy data. Noisy data refers to data where noise or unsystematic effects are present in the true values or predictions.
In particular, we introduce a new graphical method for evaluating trending, the conditional trending plot, which can represent the simultaneous evaluation of different intervals and asymmetries. 
Bootstrap methods for calculating confidence intervals for trending measures are also presented. 
In Section \ref{sec:application}, we show the results of applying our method to several practical examples from measurement, forecasting and nowcasting. 
We conclude our paper with the last section.

>>>>

@@

The main contributions of this paper are summarized below
\begin{itemize}
\item We formalize trending and show how the predicted changes can be computed to assess trending for practical relevance. 
\item We present different variants of trending measures that consider either noiseless data or data with noise and small non-informative changes.
\item We introduce the conditional trending plot, a new graphical method for assessing local trends.
\item We introduce bootstrap methods for calculating confidence intervals for the presented treding ratios.
\item We extend the concept of trending to probabilistic predictions.
\item We apply trending evaluation to three applications: Measurement, nowcasting, and forecasting data. 
\item We provide a ready-to-use code for trending detection. The code and the source code to replicate the results of our study are available in our public repository at ... .
\end{itemize}

The remainder of this paper is structured as follows. 
In Section \ref{sec:trending}, we formalize the concept of trending independently of the application domain and investigate quantification approaches for noise-free and noisy data. Noisy data refers to data where noise or unsystematic effects are present in the true values or predictions.
In particular, we introduce a new graphical method for evaluating trending, the conditional trending plot, which can represent the simultaneous evaluation of different intervals and asymmetries. 
Bootstrap methods for calculating confidence intervals for trending measures are also presented. 
In Section \ref{sec:application}, we show the results of applying our method to several practical examples from measurement, forecasting, and nowcasting. 
We conclude our paper with the last section.

### 2024-03-07 sections/03-trending.tex
Update on Overleaf.

As explained in Section~\ref{sec:introduction}, the term "trending" is used to denote the alignment of a detected true development with the development indicated by another measurement, a nowcast, or a forecast.
In this section, we want to specify trending and define different methods of measuring and visualizing trending.
In general, trending can be assessed for different lags $l$ separately, and the relevance of the considered lags has to be determined based on the question at hand.
Throughout the section, we call a measurement, nowcast, or forecast a \textit{prediction} or \textit{signal} whenever the application area is insignificant.
Before introducing measures to evaluate trending, we make remarks on the calculation of the predicted change and notation in Section~\ref{subsec:notation}.
Section~\ref{subsec:trending-basics} introduces the basics of trending and visualizes trending in a synthetic example.
In Section~\ref{subsec:trending-measures}, various methods from other disciplines and tasks are applied and examined.
The approaches are extended to account for noise and non-systematic effects in Section~\ref{subsec:trending-noise}.
Section~\ref{subsec:trending-cond-prob} introduces a new graphical method for local trending assessment and briefly introduces bootstrap methods for computing confidence intervals for the measures in the previous sections.
Trending assessment for probabilistic nowcasts and forecasts is treated in Section~\ref{subsec:probabilistic}.

@@

As noted in the introduction, the definition of the predicted change depends on the context.
Applying the computation for $y$ directly, the signal's predicted change would be the difference between the signal at time $t + l$ and the true value at $t$.
This approach lags in three practical aspects.
For nowcasting and forecasting, the true value at time $t$ might not be known with a considerable time delay.
In the nowcasting application of Section~\ref{sec:application-covid}, for example, the time delay is more than 80 days.
Until the true values are publicly available, the change could not be computed, and no actions could be taken based on it.
Thus, it is more practically relevant to use updated signal values for time $t$ as the reference value in this setting.

@@

Taking the three points together, the signal's predicted change is based on the type of signal.
To account for several signals for a target time $t$ at different issue times $\tau$, we denote them by $x_{t | \tau}$.
For nowcasting, target times $t$ are usually not considerably later than issue times $\tau$; while $t$ is usually at least $\tau$ in forecasting. For measurements, this distinction is not necessary.
Table~\ref{tab:notation} summarizes the notation for the different signal types and the computation of the predicted change.

\begin{table}
    \centering
    \begin{tabularx}{0.75\textwidth}{l X}
        \toprule
        Application & Predicted change computation \\
        \midrule
        Measurement & $(x_{t + l} - x_t)_{t=1}^T \label{eq:diffxl_measure} $\\
        Nowcasting & $
\diffxl =
\begin{cases}
(x_{t+l|t+l} -x_{t|t+l})^T_{t=1} & \text{if } y_{t} \text{ is not known at time } t, \\
(x_{t+l|t+l} -y_{t})^T_{t=1} & \text{else}.
\end{cases} \label{eq:diffxl_nowcasting}
$ \\
        Forecasting & $
\diffxl =
\begin{cases}
(x_{t+l|t} - x_{t|t})^T_{t=1} & \text{if } y_{t} \text{ is not known at time } t, \\
(x_{t+l|t} - y_{t})^T_{t=1}  & \text{else}.
\end{cases} \label{eq:diffxl_forecasting}
$\\
        \bottomrule
    \end{tabularx}
    \caption{Computation of the predicted change in the different applications. Signals $x_{t | \tau}$ refer to values issued at $\tau$ with a target time $t$. Note the different relations of issue and target time for nowcasts and forecasts. }
    \label{tab:notation}
\end{table}

@@

Although the calculation of $\diffxl$ differs in different contexts, the trending assessment is similar.
In the following, we omit the difference lag $l$ for ease of notation; $\diffx$ and $\diffy$ refer to $\diffxl$ and $\diffyl$ for a common lag $l$ as defined in Table~\ref{tab:notation}.
As noted above, a signal's trending is perfect if all the directions of change that occur are predicted correctly; that is, the sign of all elements of $\diffx$ and $\diffy$ coincide.
Perfect trending is hardly the case in practice, but a weaker requirement is needed.
Consequently, when evaluating the trending, we examine the statistical consistency of $\sign(\diffx)$ and $\sign(\diffy)$.
A simple yet insightful method is the four-quadrant plot.
For example, it is known in cardiac output measurement analysis~\parencite{Saugel2015,perrino1998intraoperative}. 
Thus, the occured changes are plotted together with the predicted changes, that is, $(\diffy_t, \diffx_t)$ for $t = 1, \dots, T$.
Thus, the x-axis of a four-quadrant plot shows the true value differences, whereas the y-axis displays the prediction data differences.
Points in the green upper right and lower left quadrants reflect a correct trending for the respective time step, whereas points in the remaining red quadrants show incorrect predicted changes.

@@

The four-quadrant plot is intuitive to interpret, and the magnitude and direction of change are shown simultaneously.
It can be extended by including information on the date of difference in the color of the point.
In Figure~\ref{fig:trending_basic_4q_sample_color}, the point colors turn from blue to green for higher time indices $t$.
The lag is always one time step.
However, four-quadrant plots become crowded for larger datasets, and sequential information on the differences is complex to assess thoroughly.
There are other visualization techniques, such as polar plots or the Bland-Altmann analysis.
However, they lack the clarity and intuition of the four-quadrant plot, without adding any more information on trending~\parencite{Saugel2015}.

@@

Analyzing the number of points in the green versus red quadrants is a standard approach in the trending assessment of measurement data \parencite{Critchley2010,Saugel2015}. 
It boils down to assessing the probability of trending $P(\diffxrv \diffyrv > 0)$, where $\diffyrv$ and $\diffxrv$ denote random variables for future incremental changes, respectively.
Note that $z_1 z_2 > 0$ if and only if $\sign(z_1) = \sign(z_2)$ ($z_1, z_2 \in \R \setminus \{ 0 \}$).
The standard estimator for $P(\diffxrv \diffyrv > 0)$ is
\begin{equation}
    \acc (\diffx, \diffy) \coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0}}{\card{T}}.\label{eq:acc}
\end{equation}
We refer to this estimator as the trending ratio of the prediction and set $\mathcal{T} = \{1, \dots, T-l\}$.
Visually, the measure computes the fraction of points in the upper right or lower left quadrant.
Such a $2 \times 2$ table of counts is called a contingency table.
It is often used in other scientific areas for evaluation, for example, in dichotomous forecasting or as a confusion matrix in classification analysis~\parencites(see, e.g., the introductions in)()[Ch. 4]{James2021}[Ch. 3]{Jolliffe2012}.
There, a wide range of other methods are usually used to analyze further characteristics of contingency tables.
Two simple measures that focus on specific areas of interest are the positive and negative trending ratios $\accp$ and $\accm$, respectively.
They are defined as
\begin{align}
    \accp (\diffx, \diffy) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt > 0}}{\sum_{t \in \mathcal{T}} \ind{\diffxt > 0}} \label{eq:accp}\\
    \accm (\diffx, \diffy) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt < 0}}{\sum_{t \in \mathcal{T}} \ind{\diffxt < 0}}\label{eq:accm}
\end{align}
In the classification context, these measures are known as positive or negative predictive value and hit rate or detection failure ratio in forecasting.
They give the probability of a correct prediction of the direction of change, given that the signal direction is positive or negative, respectively.
Therefore, they measure the trending ability of the prediction for positive and negative changes separately by cutting the considered data.
In terms of probability, the two measures give estimates of $P(\diffxrv \diffyrv > 0 | \diffxrv > 0)$ and $P(\diffxrv \diffyrv > 0 | \diffxrv < 0)$, respectively.

The use of these estimators could encourage predictors to exploit imbalances in the number of positive and negative changes. 
A large difference in $\sum_{t \in \mathcal{T}} \ind{\diffyt > 0}$ and $\sum_{t \in \mathcal{T}} \ind{\diffyt < 0}$ is unlikely in the trending setting, as $\diffy$ is obtained from differencing time series data. 
However, if the number of positive and negative $\diffy$ differs widely, the use of unbalanced-data-aware measures should be considered.
There are various adapted measures for unbalanced outcomes, for example, Cohen's $\kappa$ \parencite{Cohen1960} or those listed in \textcite[Table 3.3]{Jolliffe2012}.
Cohen's $\kappa$ is usually used to measure inter-rater agreement and takes into account the ratio of occurred agreement and the probability of agreement by chance.
Cohen's $\kappa$ reduces to rescaling the trending ratio in the case of a $2\times2$-table and balanced outcomes.

All measures can also be evaluated as a rolling estimate to detect changes in performance over time.
A rolling estimate is a sequence of estimates, each computed on a length-$w$-window of the data.
For the trending ratio, a rolling estimate with a backward-looking window is given by
\begin{equation*}
    \acc_t (\diffx, \diffy) \coloneqq \frac{\sum_{t^\star = t-w + 1}^{t} \ind{\diffxt[t^\star] \diffyt[t^\star] > 0}}{w}, \quad t = w-1, \dots, T.\label{eq:acc_rolling}
\end{equation*}
Figure~\ref{fig:trending_ratio_time_series} depicts a rolling window estimate of the trending ratio for the simulated data of Figures~\ref{fig:trending_basic_4q_sample} and~\ref{fig:trending_basic_4q_sample_color}.
Thus, the yearly course of the prediction trending ratio can be detected.
The trending ability of the signal has a strong sinus-shaped seasonality with a trending ability peak after a quarter of a year and a low point after three quarters.
This seasonal behavior cannot be distinguished in the two four-quadrant plots or in the trending ratio.

@@

In applications, signal data are often not available for all time steps, for example, due to technical problems or delays in data transfer (see the examples in Sections~\ref{sec:application-covid} and~\ref{sec:application-eda}).
We refer to time steps for which either signal or true values or both are unavailable as missing values.
Pairs of data with missing values can be excluded from the set $\mathcal{T}$ to calculate the measures. 
However, systematical missing values could lead to a biased estimate of the trending ratio, for example, if signal data is omitted in times of high change and thus uncertainty.
In this case, measures should be interpreted with caution, and
inspection of missing data should accompany applications with missing data, for example, through a visual assessment.
In the applications in Sections~\ref{sec:application-covid} and~\ref{sec:application-eda}, only a few missing values occur, and we provide information on the missing data by inspecting the missing data as lists.

@@

The above measures use only information on a point's quadrant and neglect further details on its location within the quadrant.
However, points close to the zero point have intuitively less explanatory power or are less reliable.
If there are noise or non-systematic effects present in the true values or predictions, a point's assignment to a quadrant can be driven by noise instead of a systematic trending ability of the signal.
For points with at least one small coordinate, this happens more frequently.

>>>>

As explained in Section~\ref{sec:introduction}, the term \enquote{trending} denotes the alignment of a detected true development with the development indicated by another measurement, a nowcast, or a forecast.
In this section, we want to specify trending and define different methods of measuring and visualizing trending.
In general, trending can be assessed for different lags $l$ separately, and the relevance of the considered lags has to be determined based on the question at hand.
Throughout the section, we call a measurement, nowcast, or forecast a \textit{prediction} or \textit{signal} whenever the application area is insignificant.
Before introducing measures to evaluate trending, we make remarks on calculating the predicted change and notation in Section~\ref{subsec:notation}.
Section~\ref{subsec:trending-basics} introduces trending basics and visualizes trending in a synthetic example.
In Section~\ref{subsec:trending-measures}, various methods from other disciplines and tasks are applied and examined.
The approaches are extended to account for noise and non-systematic effects in Section~\ref{subsec:trending-noise}.
Section~\ref{subsec:trending-cond-prob} introduces a new graphical method for local trending assessment and briefly introduces bootstrap methods for computing confidence intervals for the measures in the previous sections.
Trending assessment for probabilistic nowcasts and forecasts is treated in Section~\ref{subsec:probabilistic}.

@@

As noted in the introduction, the definition of the predicted change depends on the context.
Applying the computation for $y$ directly, the signal's predicted change would be the difference between the signal at time $t + l$ and the true value at $t$.
This approach lags in three practical aspects.
For nowcasting and forecasting, the true value at time $t$ might not be known with a considerable time delay.
In the nowcasting application of Section~\ref{sec:application-covid}, for example, the time delay is more than 80 days.
Until the true values are publicly available, the change cannot be computed, and no actions can be taken based on it.
Thus, using updated signal values for time $t$ as the reference value in this setting is more practically relevant.

@@

Taking the three points together, the signal's predicted change is based on the type of signal.
To account for several signals for a target time $t$ at different issue times $\tau$, we denote them by $x_{t | \tau}$.
For nowcasting, target times $t$ are usually not considerably later than issue times $\tau$, while $t$ is usually at least $\tau$ in forecasting. For measurements, this distinction is not necessary.
Table~\ref{tab:notation} summarizes the notation for the different signal types and the computation of the predicted change.

\begin{table}
    \centering
    \begin{tabularx}{0.75\textwidth}{l X}
        \toprule
        Application & Predicted change computation \\
        \midrule
        Measurement & $(x_{t + l} - x_t)_{t=1}^T \label{eq:diffxl_measure} $\\
        Nowcasting & $
\diffxl =
\begin{cases}
(x_{t+l|t+l} -x_{t|t+l})^T_{t=1} & \text{if } y_{t} \text{ is not known at time } t, \\
(x_{t+l|t+l} -y_{t})^T_{t=1} & \text{else}.
\end{cases} \label{eq:diffxl_nowcasting}
$ \\
        Forecasting & $
\diffxl =
\begin{cases}
(x_{t+l|t} - x_{t|t})^T_{t=1} & \text{if } y_{t} \text{ is not known at time } t, \\
(x_{t+l|t} - y_{t})^T_{t=1}  & \text{else}.
\end{cases} \label{eq:diffxl_forecasting}
$\\
        \bottomrule
    \end{tabularx}
    \caption{Computation of the predicted change in the different applications. Signals $x_{t | \tau}$ refer to values issued at $\tau$ with a target time $t$. Note the different relations between the issue and target time for nowcasts and forecasts. }
    \label{tab:notation}
\end{table}

@@

Although the calculation of $\diffxl$ differs in different contexts, the trending assessment is similar.
In the following, we omit the difference lag $l$ for ease of notation; $\diffx$ and $\diffy$ refer to $\diffxl$ and $\diffyl$ for a common lag $l$ as defined in Table~\ref{tab:notation}.
As noted above, a signal's trending is perfect if all the change directions are predicted correctly; that is, the sign of all elements of $\diffx$ and $\diffy$ coincide.
Perfect trending is hardly the case in practice, but a weaker requirement is needed.
Consequently, when evaluating the trending, we examine the statistical consistency of $\sign(\diffx)$ and $\sign(\diffy)$.
A simple yet insightful method is the four-quadrant plot.
For example, it is known in cardiac output measurement analysis~\parencite{Saugel2015,perrino1998intraoperative}. 
Thus, the occured changes are plotted with the predicted changes, that is, $(\diffy_t, \diffx_t)$ for $t = 1, \dots, T$.
Thus, the x-axis of a four-quadrant plot shows the true value differences, whereas the y-axis displays the prediction data differences.
Points in the green upper right and lower left quadrants reflect a correct trending for the respective time step, whereas points in the remaining red quadrants show incorrect predicted changes.

@@

The four-quadrant plot is intuitive to interpret, and the magnitude and direction of change are shown simultaneously.
It can be extended by including information on the date in the point color.
In Figure~\ref{fig:trending_basic_4q_sample_color}, the point colors turn from blue to green for higher time indices $t$.
The lag is always one time step.
However, four-quadrant plots become crowded for larger datasets, and sequential information on the differences is complex to assess thoroughly.
There are other visualization techniques, such as polar plots or the Bland-Altmann analysis.
However, they lack the four-quadrant plot's clarity and intuition, without adding more information on trending~\parencite{Saugel2015}.

@@

Analyzing the number of points in the green versus red quadrants is a standard approach in the trending assessment of measurement data \parencite{Critchley2010,Saugel2015}. 
It boils down to assessing the probability of trending $P(\diffxrv \diffyrv > 0)$, where $\diffyrv$ and $\diffxrv$ denote random variables for future incremental changes.
Note that $z_1 z_2 > 0$ if and only if $\sign(z_1) = \sign(z_2)$ ($z_1, z_2 \in \R \setminus \{ 0 \}$).
The standard estimator for $P(\diffxrv \diffyrv > 0)$ is
\begin{equation}
    \acc (\diffx, \diffy) \coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0}}{\card{T}}.\label{eq:acc}
\end{equation}
We refer to this estimator as the trending ratio of the prediction and set $\mathcal{T} = \{1, \dots, T-l\}$.
Visually, the measure computes the fraction of points in the upper right or lower left quadrant.
Such a $2 \times 2$ table of counts is called a contingency table.
It is often used in other scientific areas for evaluation, for example, in dichotomous forecasting or as a confusion matrix in classification analysis~\parencites(see, e.g., the introductions in)()[Ch. 4]{James2021}[Ch. 3]{Jolliffe2012}.
There, a wide range of other methods are usually used to analyze further characteristics of contingency tables.
Two simple measures that focus on specific areas of interest are the positive and negative trending ratios $\accp$ and $\accm$, respectively.
They are defined as
\begin{align}
    \accp (\diffx, \diffy) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt > 0}}{\sum_{t \in \mathcal{T}} \ind{\diffxt > 0}} \label{eq:accp}\\
    \accm (\diffx, \diffy) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt < 0}}{\sum_{t \in \mathcal{T}} \ind{\diffxt < 0}}\label{eq:accm}
\end{align}
In the classification context, these measures are known as positive or negative predictive value and hit rate or detection failure ratio in forecasting.
They give the probability of a correct prediction of the direction of change, given that the signal direction is positive or negative, respectively.
Therefore, they measure the trending ability of the prediction for positive and negative changes separately by cutting the considered data.
In terms of probability, the two measures give estimates of $P(\diffxrv \diffyrv > 0 | \diffxrv > 0)$ and $P(\diffxrv \diffyrv > 0 | \diffxrv < 0)$, respectively.

Using these estimators could encourage predictors to exploit imbalances in the number of positive and negative changes. 
A large difference in $\sum_{t \in \mathcal{T}} \ind{\diffyt > 0}$ and $\sum_{t \in \mathcal{T}} \ind{\diffyt < 0}$ is unlikely in the trending setting, as $\diffy$ is obtained from differencing time series data. 
However, if the number of positive and negative $\diffy$ differs widely, the use of unbalanced-data-aware measures should be considered.
There are various adapted measures for unbalanced outcomes, for example, Cohen's $\kappa$ \parencite{Cohen1960} or those listed in \textcite[Table 3.3]{Jolliffe2012}.
Cohen's $\kappa$ is usually used to measure inter-rater agreement and considers the ratio of occurred agreement and the probability of agreement by chance.
Cohen's $\kappa$ reduces to rescaling the trending ratio for a $2\times2$-table and balanced outcomes.

All measures can also be evaluated as a rolling estimate to detect changes in performance over time.
A rolling estimate is a sequence of estimates, each computed on a length-$w$-window of the data.
For the trending ratio, a rolling estimate with a backward-looking window is given by
\begin{equation*}
    \acc_t (\diffx, \diffy) \coloneqq \frac{\sum_{t^\star = t-w + 1}^{t} \ind{\diffxt[t^\star] \diffyt[t^\star] > 0}}{w}, \quad t = w-1, \dots, T.\label{eq:acc_rolling}
\end{equation*}
Figure~\ref{fig:trending_ratio_time_series} depicts a rolling window estimate of the trending ratio for the simulated data of Figures~\ref{fig:trending_basic_4q_sample} and~\ref{fig:trending_basic_4q_sample_color}.
Thus, the yearly course of the prediction trending ratio can be detected.
The trending ability of the signal has a strong sinus-shaped seasonality with a trending ability peak after a quarter of a year and a low point after three quarters.
This seasonal behavior cannot be distinguished in the two four-quadrant plots or the trending ratio.

@@

In applications, signal data are often not available for all time steps, for example, due to technical problems or delays in data transfer (see the examples in Sections~\ref{sec:application-covid} and~\ref{sec:application-eda}).
We refer to time steps for which either signal or true values or both are unavailable as missing values.
Pairs of data with missing values can be excluded from the set $\mathcal{T}$ to calculate the measures. 
However, systematical missing values could lead to a biased estimate of the trending ratio, for example, if signal data is omitted in times of high change and thus uncertainty.
In this case, measures should be interpreted cautiously, and
inspection of missing data should accompany applications with missing data, for example, through a visual assessment.
In the applications in Sections~\ref{sec:application-covid} and~\ref{sec:application-eda}, only a few missing values occur, and we provide information on the missing data by inspecting the missing data as lists.

@@

The above measures use only information on a point's quadrant and neglect further details on its location within the quadrant.
However, points close to the zero point have intuitively less explanatory power or are less reliable.
Suppose there are noise or non-systematic effects present in the true values or predictions. In that case, a point's assignment to a quadrant can be driven by noise instead of a systematic trending ability of the signal.
This happens more frequently for points with at least one small coordinate.
